Distinct smell and taste disorder phenotype of post-acute COVID-19 sequelae

### 2023-05-30 paper/markdown/supplementary_material.Rmd
Figures, Tables, Supplementary Material

library(bookdown)

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, 
                      dev = "png", dpi = 600, 
                      tab.cap.style = 'Table Heading')

@@

A total of self-reported `r length(globals$hact_symptoms)` symptoms were recorded in the survey study cohorts (__Supplementary Figure S\@ref(fig:fig-hact-freq)__ and __Supplementary Table S\@ref(tab:tab-hact-vars)__). The symptom duration was coded as follows: absent: 0 days, 1 - 3 days: 3 days, up to 1 week: 7 days, up to 2 weeks: 14 days, up to 4 weeks: 28 days, up to or greater than 3 months: 3 months. Acute symptoms were defined as complaints present during the first 14 days after clinical onset of COVID-19.

In the observatory CovILD cohort, a total of self-reported `r length(globals$covild_symptoms)` symptoms (reduced physical performance, olfactory dysfunction, dyspnea, sleep problems, cough, fever, night sweating, gastrointestinal symptoms) were recorded with a standardized questionnaire at each of 60-, 100-, 180- and 360-day post COVID-19 follow-up (__Supplementary Figure S\@ref(fig:fig-covild-freq)__ and __Supplementary Table S\@ref(tab:tab-covild-vars)__). Acute COVID-19 symptoms were assessed retrospectively at the 60-day follow-up [@Sonnweber2020; @Sonnweber2022].

@@

Self-perceived complete recovery, rehabilitation need and new medication since COVID-19 at the time of study participation were surveyed as single yes/no items. Percentage of physical performance loss as compared with the time before COVID-19 was rated with a 0 - 100% scale [@Hufner2022; @Sahanic2021]. Quality of life impairment (QoL) and overall mental health impairment (OMH) were rated with a 4 item Likert scale each (possible answers: “excellent,” “good,” “fair,” “poor,” scored: 0, 1, 2, and 3) [@Hufner2022; @Sahanic2021]. Anxiety (ANX) and depression (DPR) were assessed with the Patient Health Questionnaire (PHQ-4) [@Hufner2022; @Sahanic2021; @Lowe2010]. Psychosocial stress was scored with a modified 7 item PHQ stress module as described before [@Hufner2022; @Sahanic2021; @Grafe2004].

### Rating of olfactory dysfunction with sniffing stick test

Objective olfactory dysfunction at the 100-day and 360-day follow-up in the CovILD study participants was investigated with the 16-item sniffing stick test as described [@Rass2021]. Clinically relevant olfactory dysfunction was defined as $\leq$ 12 correct answers [@Rass2021; @Hummel2007]. In the analysis, participants with the complete answers concerning self-reported olfactory dysfunction and complete test results were included.

## Statistical analysis
### Data transformation, descriptive statistic

Data transformation and statistical analysis was done with R version 4.2.0 with _tidyverse_ data science environment [@Wickham2019]. Analysis results were visualized with _ggplot2_ [@Wickham2016], _cowplot_ [@Wilke2019] as well as in-house developed _ExDA_ (https://github.com/PiotrTymoszuk/ExDA) and _figur_ (https://github.com/PiotrTymoszuk/figur) packages.

Descriptive statistics including median with interquartile ranges and frequency of complete answers for numeric and categorical variables were calculated with base R functions and _ExDA_ package.

### Statistical hypothesis testing

Since multiple study variables were non-normally distributed as assessed by Shapiro-Wilk test and visual assessment of their distribution (quantile-quantile plots), statistical significance for differences in outcome numeric variables were assessed with Mann-Whitney U test with r effect size statistic (two groups) or Kruskal-Wallis test with $\eta^2$ effect size statistic. Differences in frequency distribution for categorical outcome variables were assessed by $\chi^2$ test with Cramer V effect size statistic. Inter-rater assessment of self-reported and sniffing test olfactory dysfunction was accomplished with Cohen's $\kappa$ statistic [@Fleiss1969]. $\kappa$ significance ($\kappa\neq0$) was estimated with Wald Z test. P values were adjusted for multiple testing with Benjamini-Hochberg method [@Benjamini1995] separately for each analysis task and cohort. 
Packages _rstatix_ [@Kassambara2021], _vcd_ [@Meyer2021] and _Exda_ (https://github.com/PiotrTymoszuk/ExDA) were used for statistical hypothesis testing.

### Modeling of symptom recovery kinetic

To model recovery kinetics for binary symptom variables (0: absent, 1: present), second-order mixed-effect logistic (categorical features) modeling was applied (packages: _lme4_, _lmerTest_ and development package _kinet_ [https://github.com/PiotrTymoszuk/kinet]) [@Bates2015; @Kuznetsova2017; @Box1978]. Each model followed the general formula:

$$Response \sim time + time^2 + (1|individual)$$

where $(1|individual)$ indicates the random effect of the individual and $time$ and $time^2$ indicate the first- and second-order time effect terms. The first-order term estimate was interpreted as a measure of the recovery speed and the second-order term estimate was used to assess the plateau/rebound effect. Significance of the accuracy gain of the full second-order model compared with the nested null model was determined by likelihood ratio test (LRT) versus the nested first-order and null models, respectively. Likelihood ratio $\lambda$ statistic (full versus null model) was used as an effect size measure. 
Individuals from the survey study or the CovILD cohort with the complete longitudinal symptom record were included in the kinetic modeling tasks. Results of the kinetic modeling were adjusted for multiple comparisons with Benjamini-Hochberg method [@Benjamini1995].

@@

To assess co-occurrence or exclusivity of symptoms, simple matching distances between manifestations during the first 14 days, at $\geq$ 28 days and at $\geq$ 3 months after clinical onset in the survey study cohorts were calculated (package _scrime_ and development package _clustTools_ [https://github.com/PiotrTymoszuk/clustTools]) [@Boriah2008; @Schwender2018]. Subsequently, the distance matrix was subjected to multi-dimensional scaling (MDS, k = 2 dimensions, package _stats_, function _cmdscale()_). Association of specific symptoms was assessed by visual analysis of MDS coordinate plots.

@@

Frequent combinations of symptoms during the first 14 days, at $\geq$ 28 days and at $\geq$ 3 months after clinical onset in the survey study cohorts were identified with the apriori algorithm (package _arules_) [@Agrawal1993; @Hahsler2005] with the minimal support cutoff of 0.15, 2 - 10 item transaction length, confidence > 0.6 and lift > 2. The support statistic were used to estimate the symptom combination frequency. The confidence value was treated as an estimate of conditional probability of the symptom co-occurrence. The lift statistic was interpreted as a measure of the symptom dependence (lift = 1, symptoms are independent).

@@

COVID-19 recovery clusters of the training Autria (AT) survey cohort participants in respect to symptom-specific recovery times (__Figure 1A__) were defined with the PAM (partitioning around medoids) algorithm and Euclidean distance statistic (packages _cluster_, _philentropy_ and development package _clustTools_ [https://github.com/PiotrTymoszuk/clustTools]) [@Drost2018; @Schubert2019]. The set of participants with the complete clustering variable set was included in the analysis. The symptom recovery times were not subjected to any type of pre-processing. The choice of the clustering procedure was motivated by the analysis of the clustering variance (ratio of the total between-cluster to total sum of squares) and clustering structure stability in 10-fold cross-validation (metric: rate of correct cluster assignment, cluster assignment predicted by k = 5 nearest neighbors label propagation algorithm, package _clustTools_) [@Lange2004; @Leng2014] for several clustering algorithms as presented in __Supplementary Figure S\@ref(fig:fig-clust-devel)A__. The optimal number of clusters was determined by the bend of the total within-cluster sum of squares curve (__Supplementary Figure S\@ref(fig:fig-clust-devel)B__, package _factoextra_) [@Kassambara2020]. 
Permutation importance of specific clustering variables was investigated by calculating difference in clustering variance (ratio of total between-cluster sum of squares to total sum of squares) between the initial clustering object and the clustering object with the given variable reshuffled at random (package _clustTools_). 
Assignment of the Italy survey cohort participants to the recovery clusters was accomplished with k-nearest neighbors label propagation algorithm (k = 5) [@Leng2014]. The clustering efficacy in the training AT cohort and the test IT cohort measured by clustering variance statistic defined above was similar (AT: `r clustTools::var(part_clust$clust_north)$frac_var %>% signif(2)`, IT: `r clustTools::var(part_clust$clust_south)$frac_var %>% signif(2)`).

@@

The raw data files will be made available upon request. The entire analysis pipeline was published at https://github.com/PiotrTymoszuk/hyposmia_analsis_pipeline.

@@

```{r tab-hact-vars, tab.cap = 'Survey study variables.'}

suppl_tbl$hact_vars %>% 
  select(- Stratification) %>% 
  flextable %>% 
  width(1, width = 3.7, unit = 'cm') %>% 
  width(2, width = 3.9, unit = 'cm') %>% 
  width(3, width = 1.8, unit = 'cm')  %>% 
  width(4, width = 8.7, unit = 'cm') %>% 
  theme_vanilla

@@

```{r tab-covild-vars, tab.cap = 'CovILD study variables.'}

suppl_tbl$covild_vars %>% 
  flextable %>% 
  width(1, width = 4.5, unit = 'cm') %>% 
  width(2, width = 4.5, unit = 'cm') %>% 
  width(3, width = 2, unit = 'cm')  %>% 
  theme_vanilla

@@

```{r tab-clust-base-at, tab.cap = 'Demographic and baseline clinical characteristic at the COVID-19 onset of the survey study participants assigned to the recovery clusters, Austria (AT) cohort.'}

suppl_tbl$baseline_clusters_north %>% 
  flextable %>% 
  width(1, width = 3.5, unit = 'cm') %>% 
  width(2:4, width = 3.7, unit = 'cm') %>% 
  width(5, width = 2.6, unit = 'cm') %>% 
  width(6, width = 2.2, unit = 'cm') %>% 
  footnote(1, 5:6, value = as_paragraph('Categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic. Numeric variables: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic. P values corrected form multiple testing with Benjamini-Hochberg method.'), ref_symbols = 'a', part = 'header') %>% 
  footnote(3, 1, value = as_paragraph('BMI: body mass index, overweight > 25 kg/m\u00B2, obesity > 30 kg/m\u00B2,'), ref_symbols = 'b', part = 'body') %>% 
  footnote(15, 1, value = as_paragraph('Frequent respiratory infections, > 2 per year.'), ref_symbols = 'c', part = 'body') %>% 
  footnote(16, 1, value = as_paragraph('Frequent bacterial infections with antibiotic therapy, > 2 per year.'), ref_symbols = 'd', part = 'body') %>% 
  theme_vanilla

@@

```{r tab-clust-base-it, tab.cap = 'Demographic and baseline clinical characteristic at the COVID-19 onset of the survey study participants assigned to the recovery clusters, Italy (IT) cohort.'}

suppl_tbl$baseline_clusters_south %>% 
  flextable %>% 
  width(1, width = 3.5, unit = 'cm') %>% 
  width(2:4, width = 3.7, unit = 'cm') %>% 
  width(5, width = 2.6, unit = 'cm') %>% 
  width(6, width = 2.2, unit = 'cm') %>% 
  footnote(1, 5:6, value = as_paragraph('Categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic. Numeric variables: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic. P values corrected form multiple testing with Benjamini-Hochberg method.'), ref_symbols = 'a', part = 'header') %>% 
  footnote(3, 1, value = as_paragraph('BMI: body mass index, overweight > 25 kg/m\u00B2, obesity > 30 kg/m\u00B2,'), ref_symbols = 'b', part = 'body') %>% 
  footnote(15, 1, value = as_paragraph('Frequent respiratory infections, > 2 per year.'), ref_symbols = 'c', part = 'body') %>% 
  footnote(16, 1, value = as_paragraph('Frequent bacterial infections with antibiotic therapy, > 2 per year.'), ref_symbols = 'd', part = 'body') %>% 
  theme_vanilla

@@

```{r tab-clust-reco-at, tab.cap = 'COVID-19 course and recovery in the survey study participants assigned to the recovery clusters, Austria (AT) cohort.'}

suppl_tbl$cov_clusters_north %>% 
  flextable %>% 
  width(1, width = 3.5, unit = 'cm') %>% 
  width(2:4, width = 3.7, unit = 'cm') %>% 
  width(5, width = 2.6, unit = 'cm') %>% 
  width(6, width = 2.2, unit = 'cm') %>% 
  footnote(1, 5:6, value = as_paragraph('Categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic. Numeric variables: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic. P values corrected form multiple testing with Benjamini-Hochberg method.'), ref_symbols = 'a', part = 'header') %>% 
  footnote(4, 1, value = as_paragraph('Self-reported incomplete recovery.'), ref_symbols = 'd', part = 'body') %>% 
  footnote(8, 1, value = as_paragraph('ANX: anxiety.'), ref_symbols = 'e', part = 'body') %>% 
  footnote(9, 1, value = as_paragraph('DPR: depression.'), ref_symbols = 'f', part = 'body') %>% 
  footnote(11, 1, value = as_paragraph('OMH score: overall mental health impairment score'), ref_symbols = 'g', part = 'body') %>% 
  footnote(12, 1, value = as_paragraph('QoL score: quality of life impairment score'), ref_symbols = 'h', part = 'body') %>%
  theme_vanilla

@@

```{r tab-clust-reco-it, tab.cap = 'COVID-19 course and recovery in the survey study participants assigned to the recovery clusters, Italy (IT) cohort.'}

suppl_tbl$cov_clusters_south %>% 
  flextable %>% 
  width(1, width = 3.5, unit = 'cm') %>% 
  width(2:4, width = 3.7, unit = 'cm') %>% 
  width(5, width = 2.6, unit = 'cm') %>% 
  width(6, width = 2.2, unit = 'cm') %>% 
 footnote(1, 5:6, value = as_paragraph('Categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic. Numeric variables: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic. P values corrected form multiple testing with Benjamini-Hochberg method.'), ref_symbols = 'a', part = 'header') %>% 
  footnote(4, 1, value = as_paragraph('Self-reported incomplete recovery.'), ref_symbols = 'd', part = 'body') %>% 
  footnote(8, 1, value = as_paragraph('ANX: anxiety.'), ref_symbols = 'e', part = 'body') %>% 
  footnote(9, 1, value = as_paragraph('DPR: depression.'), ref_symbols = 'f', part = 'body') %>% 
  footnote(11, 1, value = as_paragraph('OMH score: overall mental health impairment score'), ref_symbols = 'g', part = 'body') %>% 
  footnote(12, 1, value = as_paragraph('QoL score: quality of life impairment score'), ref_symbols = 'h', part = 'body') %>%
  theme_vanilla

>>>>

library(bookdown)
library(Cairo)

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, 
                      dev = "CairoPNG", dpi = 600, 
                      tab.cap.style = 'Table Heading')

@@

A total of `r length(globals$hact_symptoms)` self-reported symptoms were recorded in the survey study cohorts 
(__Supplementary Figure S\@ref(fig:fig-hact-freq)__ and __Supplementary Table S\@ref(tab:tab-hact-vars)__).
The symptom duration was coded as follows: absent: 0 days, 1 - 3 days: 3 days, up to 1 week: 7 days, up to 2 weeks: 14 days, up to 4 weeks: 28 days, up to or greater than 3 months: 3 months. 
Acute symptoms were defined as complaints present during the first 14 days after clinical onset of COVID-19.

In the observatory CovILD cohort, a total of `r length(globals$covild_symptoms)` self-reported symptoms 
(reduced physical performance, olfactory dysfunction, dyspnea, sleep problems, cough, fever, night sweating, gastrointestinal symptoms) were recorded with a standardized questionnaire at each of 60-, 100-, 180- and 360-day post COVID-19 follow-up (__Supplementary Figure S\@ref(fig:fig-covild-freq)__ and __Supplementary Table S\@ref(tab:tab-covild-vars)__). 
Acute COVID-19 symptoms were assessed retrospectively at the 60-day follow-up [@Sonnweber2020; @Sonnweber2022].

@@

Self-perceived complete recovery, rehabilitation need and new medication since COVID-19 at the time of study participation were surveyed as single yes/no items. 
Percentage of physical performance loss as compared with the time before COVID-19 was rated with a 0 - 100% scale [@Hufner2022; @Sahanic2021]. 
Quality of life impairment (QoL) and overall mental health impairment (OMH) were rated with a 4 item Likert scale each (possible answers: “excellent,” “good,” “fair,” “poor,” scored: 0, 1, 2, and 3) [@Hufner2022; @Sahanic2021]. 
Anxiety (ANX) and depression (DPR) were assessed with the Patient Health Questionnaire (PHQ-4) [@Hufner2022; @Sahanic2021; @Lowe2010]. 
Mental stress was scored with a modified 7 item PHQ stress module as described before [@Hufner2022; @Sahanic2021; @Grafe2004].

### Rating of olfactory dysfunction with Sniffin' Stick Test

Objective olfactory dysfunction at the 100-day and 360-day follow-up in the CovILD study participants was investigated with the 16-item Sniffn' Stick Identification Test as described [@Rass2021; @Whitcroft2016; @Masala2022; @Damm2014]. 
In brief, the nasal chemosensory performance was investigated using pen-like odor-dispensing devices for odor identification of 16 common odorants (multiple forced‐choice from four verbal items per test odorant).
Clinically relevant olfactory dysfunction was defined as < 13 correct answers (points) [@Rass2021; @Hummel2007]. 
In the analysis, participants with the complete answers concerning self-reported olfactory dysfunction and complete test results were included.

## Data analysis

### Software

Analysis was done with R version 4.2.3 (The R Project for Statistical Computing).

Import of raw study data was accomplished with the packages _foreign_ [@RCoreTeam] and _readxl_ [@Wickham2022]. 
Tabular data was handled with the _tidyverse_ package bundle [@Wickham2019], and the _rlang_ [@Henry2022] and [_trafo_](https://github.com/PiotrTymoszuk/trafo) packages. 
Search and transformation of text data was done with _stringi_ [@Gagolewski2021]. 
For exploratory data analysis, descriptive statistic and statistical hypothesis testing, packages _rstatix_ [@Kassambara2021], _vcd_ [@Meyer2021], _rcompanion_ [@Mangiafico2022] and [_ExDA_](https://github.com/PiotrTymoszuk/ExDA) were used. 
For modeling of symptom recovery, packages _lme_ [@Bates2015], _lmerTest_ [@Kuznetsova2017] and [_kinet_](https://github.com/PiotrTymoszuk/kinet) were employed. 
Multi-dimensional scaling was performed with the base R's _stats_ package and [_clustTools_](https://github.com/PiotrTymoszuk/clustTools). 
Apriori analysis was done with _arules_ [@Hahsler2005]. 
Clustering analysis was performed with the packages _cluster_ [@Schubert2019], _philentropy_ [@Drost2018], _scrime_ [@Schwender2018], _factoextra_ [@Kassambara2020] and [_clustTools_](https://github.com/PiotrTymoszuk/clustTools). 
Inter-rater analysis and receiver operating characteristic (ROC) were done with the packages _caret_ [@Kuhn2008] and _vcd_ [@Meyer2021]. 
Parallelization of analysis tasks was accomplished with the package _furrr_ package [@Vaughan2022].

Analysis results were visualized with _ggplot_ [@Wickham2016] (bar, scatter and bubble plots, heat maps of confusion matrices), [_ExDA_](https://github.com/PiotrTymoszuk/ExDA) (ellipse plots, radar, ribbon, stack and violin plots), [_kinet_](https://github.com/PiotrTymoszuk/kinet) (kinetic of symptom resolution), 
_plotROC_ [@Sachs2017] (ROC plots), [_clustTools_](https://github.com/PiotrTymoszuk/clustTools) (multi-dimensional scaling projection/scatter plots, heat maps of clustering features, within-cluster sum of squares). 
HTML elements within the plots were handled with tools provided by the package _ggtext_ [@Wilke2022].

Figures were generated with _cowplot_ [@Wilke2019]. 
Tables were rendered with the _flextable_ package [@Gohel2022]. 
Supplementary Material and parts of the manuscript were written in the _rmarkdown_ environment [@Allaire2022] and were rendered as Word documents with the _knitr_ [@Xie2022] and _bookdown_ [@Xie2016] packages. 
Management of figures and tables within the _rmarkdown_ documents was accomplished with the development package [_figur_](https://github.com/PiotrTymoszuk/figur).

### Data import and formatting

Data of the survey study [@Sahanic2021; @Hufner2022] were imported from SPSS files with the _foreign_ package (function `read.spss()`). 
The CovILD study data set [@Sahanic2023] was imported from an Excel file with the function `read_xlsx()` (package _readxl_). 
Formatting of the raw data sets was done with in-house developed scripts available from the [project's GitHub repository](https://github.com/PiotrTymoszuk/hyposmia_analysis_pipeline).

### Estimation of minimal sample size for clustering analysis

To assess the minimal sample size of the survey study required for reliable clustering analysis, random subsets of the pooled Austria and Italy cohort COVID-19 symptom data set with differing observation numbers (50, 100, 200, 300, 400, 500, 600, 700, 800, 900 observations, 20 random subsets per sample size) were generated and their clustering tendency was assessed with Hopkins statistic (H). 
Possible H values span from 0 to 1, where H = 0 indicates an ideal uniform distribution and H = 1 suggests a highly clustered data. 
Beginning from the sample size of n = 400, no improvement of the clustering tendency (plateau) could be observed. 
This suggest n = 400 as minimal sample size required for reproducible clustering analysis results (__Supplementary Figure S\@ref(fig:fig-sample-size)__). 
Given the size of the Austria and Italy cohorts 
(n = `r nrow(part_clust$analysis_tbl$north)` and n = `r nrow(part_clust$analysis_tbl$south)`, respectively) and their H values being very close to the H plateau value in the random subset analysis 
(Austria: H = `r signif(eda_size$tendency_cohorts$north$hopkins_stat, 2)`, 
Italy: H = `r signif(eda_size$tendency_cohorts$south$hopkins_stat, 2)`), we inferred that the size of single study cohorts was sufficient for reliable and reproducible clustering analysis and hence abstained from pooling the survey study cohorts.

### Descriptive statistic and effect size

If not indicated otherwise, numeric values are presented in the manuscript text and tables as medians with interquartile ranges (IQR) and ranges. 
Qualitative variables are presented as percentages and counts of their categories within the complete observation set. 
Descriptive statistic values were computed with the function `explore()` from the [_ExDA_](https://github.com/PiotrTymoszuk/ExDA) package.

Effects sizes were assessed following the scheme proposed by Cohen [@Cohen2013]. 
Effect size for comparison of categorical variables was measured with Cramer's V statistic (weak: < 0.3, moderate: 0.3 - 0.5, large effect: $\geq$ 0.5). 
Effect size for two-group comparisons of numeric variables was assessed with r statistic (weak: < 0.3, moderate: 0.3 - 0.5, large effect: $\geq$ 0.5). 
Effect size for multi-group comparison of numeric variables was assessed with $\eta^2$ (weak: < 0.13, moderate: 0.13 - 0.26, large effect: $\geq$ 0.26). 
Effect size for comparison of paired proportions (2 $\times$ 2 contingency table) was assessed with Cohen's g (weak: < 0.15, moderate: 0.15 - 0.25, large effect: $\geq$ 0.25). 
Effect size for comparison of study participant-matched proportions (multi-dimensional contingency table) was assessed with Kendall's W (weak: < 0.3, moderate: 0.3 - 0.5, large effect: $\geq$ 0.5). 
Effect size for inter-rate reliability was measures with Cohen's $\kappa$ statistic was interpreted as follows: 0.01 – 0.20 as none to slight, 0.21 – 0.40 as fair, 0.41 – 0.60 as moderate, 0.61 – 0.80 as substantial, and 0.81 – 1.00 as strong [@McHugh2012]. 
Effect sizes were computed with the function `compare_variables()` from the development package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA) employing tools introduced by the _rstatix_ and _vcd_ packages (Cramer's V, r, $\eta^2$, Cohen's $\kappa$, Cohen's g) or with the function `cohenG()` from the package _rcompanion_ [@Mangiafico2022].

### Statistical hypothesis testing and inter-rater reliability

Since multiple study variables were non-normally distributed as assessed by Shapiro-Wilk test and visual assessment of their distribution (quantile - quantile plots), statistical significance for differences in outcome numeric variables were assessed with Mann-Whitney U test with r effect size statistic (two groups, independent data), paired Wilcoxon test with r effect size statistic (two groups, paired data) or Kruskal-Wallis test with $\eta^2$ effect size statistic. 
Differences in frequency distribution for categorical outcome variables were assessed by $\chi^2$ test with Cramer V effect size statistic. 
Significance of differences of paired proportions (2 $\times$ 2 contingency table) was determined with McNemar test with Cohen's g effect size statistic. 
Mann-Whitney, Kruskal-Wallis, $\chi^2$ tests were accomplished with the function `compare_variables()` (package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA)) employing algorithms from the package _rstatix_ [@Kassambara2021]. 
McNemar tests and Cohen's g were calculated with the functions `mcnemar_test()` (package _rstatix_) [@Kassambara2021] and `cohenG()` (package _rcompanion_) [@Mangiafico2022], respectively. 
Significance of symptom resolution (longitudinal, participant-matched binary data) was determined by Cochran's Q test with Kendall's W effect size statistic (function `cochran_qtest()`, package _rstatix_) [@Serlin1982; @Kassambara2021].

Inter-rater assessment of self-reported and Sniffin Test olfactory dysfunction was accomplished with Cohen's $\kappa$ statistic [@Fleiss1969; @McHugh2012]. 
$\kappa$ significance ($\kappa\neq0$) was estimated with Wald Z test (function `correlate_variable()`, package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA) employing the _vcd_ package) [@Meyer2021]. 
Additionally, reliability of detection of objective olfactory dysfunction (Sniffin' Stick Test) was additionally assessed by ROC analysis. 
Overall accuracy, sensitivity and specificity was computed with the functions `defaultSummary()` and `twoClassSummary()` from the package _caret_ [@Kuhn2008].

P values were adjusted for multiple testing with Benjamini-Hochberg method [@Benjamini1995] separately for each analysis task and cohort. 
Effects with p < 0.05 were considered significant.

@@

To assess co-occurrence or exclusivity of symptoms, simple matching distances between manifestations during the first 14 days, at 28 days and at 3 months after clinical onset in the survey study cohorts were calculated (function `calculate_dist()`, package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools) employing tools from packages _scrime_ and _philentropy_) [@Boriah2008; @Schwender2018; @Drost2018]. 
Subsequently, the distance matrix was subjected to multi-dimensional scaling (MDS, k = 2 dimensions, package _stats_, function `cmdscale()`). 
Association of specific symptoms was assessed by visual analysis of MDS coordinate plots.

@@

Frequent combinations of symptoms during the first 14 days, at 28 days and at 3 months after clinical onset in the survey study cohorts were identified with the apriori algorithm (function `apriori()`, package _arules_) [@Agrawal1993; @Hahsler2005] with the minimal support cutoff of 0.15, 2 - 10 item transaction length, confidence > 0.6 and lift > 2. 
The support statistic were used to estimate the symptom combination frequency. 
The confidence value was treated as an estimate of conditional probability of the symptom co-occurrence. 
The lift statistic was interpreted as a measure of the symptom dependence (lift = 1, symptoms are independent). 
Frequency of symptom combination in the survey study cohorts and percentage of co-occurrence of the symptoms within the given symptom combination were displayed in bobble plots.

@@

COVID-19 recovery clusters of the training Austria survey cohort participants in respect to symptom-specific recovery times (__Figure 1A__) were defined with the PAM (partitioning around medoids) algorithm and Euclidean distance statistic [@Drost2018; @Schubert2019]. 
The set of participants with the complete clustering variable set (COVID-19 symptom recovery times) was included in the analysis. 
The symptom recovery times were not subjected to any type of pre-processing. 
The clustering objects were generated with the function `kcluster(clust_fun = 'pam')` from the package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools).

The choice of the clustering procedure was motivated by the analysis of the fraction of explained clustering variance (ratio of the total between-cluster to total sum of squares) and clustering structure stability in 10-fold cross-validation (metric: rate of correct cluster assignment, cluster assignment predicted by inverse distance-weighted 7-nearest neighbors label propagation algorithm) [@Lange2004; @Leng2014] for several clustering algorithms as presented in __Supplementary Figure S\@ref(fig:fig-clust-devel)A__. 
The fractions of explained clustering variance and cross-validated cluster assignment accuracy were calculated with the methods `var()` and `cv()` from the package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools).

The optimal number of clusters was determined by the bend of the total within-cluster sum of squares curve 
(__Supplementary Figure S\@ref(fig:fig-clust-devel)B__, method `plot()`, package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools), employing the genuine _factoextra_ algorithm) [@Kassambara2020]. 
Permutation importance of specific clustering variables was investigated by calculating difference in clustering variance (ratio of total between-cluster sum of squares to total sum of squares) between the initial clustering object and the clustering object with the given variable reshuffled at random (function `impact()`, package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools)). 
The permutation importance statistics were computed for 20 random permutations of each clustering variable.

Assignment of the Italy survey cohort participants to the recovery clusters was accomplished with the inverse distance-weighted 7-nearest neighbors label propagation classifier [@Leng2014]. 
The clustering efficacy in the training Austria cohort and the test Italy cohort measured by clustering variance statistic defined above was similar 
(Austria: `r clustTools::var(part_clust$clust_north)$frac_var %>% signif(2)`, 
Italy: `r clustTools::var(part_clust$clust_south)$frac_var %>% signif(2)`), which indicate good reproducibility of the clustering structure developed in the training Austria cohort.

@@

The raw data files will be made available upon request. The entire analysis pipeline was published at https://github.com/PiotrTymoszuk/hyposmia_analysis_pipeline.

@@

```{r tab-hact-vars, tab.cap = 'Survey study variables. The table is available as a supplementary Excel sheet.'}

suppl_tbl$hact_vars[1, 1] %>% 
  mutate(`Variable name` = '') %>% 
  set_names(' ') %>% 
  flextable

@@

```{r tab-covild-vars, tab.cap = 'CovILD study variables. The table is available as a supplementary Excel sheet.'}

suppl_tbl$covild_vars[1, 1] %>% 
  mutate(`Variable name` = '') %>% 
  set_names(' ') %>% 
  flextable

@@

```{r tab-hact-kinetic, tab.cap = 'Results of statistical hypothesis testing for significant recovery of the most frequent COVID-19 symptoms in the Austria (AT) and Italy cohort (IT) of the survey study.'}

flextable::flextable(suppl_tbl$hact_kinetic) %>% 
  width(1, width = 2.5, unit = 'cm') %>% 
  width(2, width = 4.5, unit = 'cm') %>% 
  width(3:4, width = 3, unit = 'cm') %>% 
  merge_v(1) %>% 
  footnote(1, 2, 
           value = as_paragraph('OD: self-reported olfactory dysfunction; Dim. appetite: diminished appetite.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 3:4, 
           value = as_paragraph("Cochran's Q test with Kendall's W effect size statistic. P values corrected for multiple testing with Benjamini-Hochberg method."), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  theme_vanilla

@@

```{r tab-covild-kinetic, tab.cap = 'Results of statistical hypothesis testing for significant recovery of COVID-19 symptoms in COVID-19 severity strata of the CovILD study.'}

flextable::flextable(suppl_tbl$covild_kinetic) %>% 
  width(1, width = 3, unit = 'cm') %>% 
  width(2, width = 4.5, unit = 'cm') %>% 
  width(3:4, width = 3, unit = 'cm') %>% 
  merge_v(1) %>% 
  footnote(1, 2, 
           value = as_paragraph('OD: self-reported olfactory dysfunction.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 3:4, 
           value = as_paragraph("Cochran's Q test with Kendall's W effect size statistic. P values corrected for multiple testing with Benjamini-Hochberg method."), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  theme_vanilla

@@

```{r tab-od-kinetic, tab.cap = "Results of the Sniffin' Stick Test in the CovILD study subset with the complete longitudinal follow-up data. Numeric variables are presented as medians with interquartile ranges (IQR) and ranges. Categorical variables are presented as percentages and counts within the complete observation set."}

flextable::flextable(suppl_tbl$od_kinetic) %>% 
  width(1, width = 4, unit = 'cm') %>% 
  width(2:3, width = 4.5, unit = 'cm') %>% 
  width(4, width = 2.7, unit = 'cm') %>% 
  width(5, width = 2.5, unit = 'cm') %>% 
  footnote(1, 1, 
           value = as_paragraph('OD: olfactory dysfunction.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 3:4, 
           value = as_paragraph("Categorical variables: McNemar test with Cohen's g effect size statistic. Numeric variables: paired Wilcoxon test with r effect size statistic. P values corrected for multiple testing with Benjamini-Hochberg method."), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  theme_vanilla

@@

```{r tab-clust-base-at, tab.cap = 'Demographic and baseline clinical characteristic at the COVID-19 onset of the survey study participants assigned to the recovery clusters, Austria (AT) cohort. Numeric variables are presented as medians with interquartile ranges (IQR) and ranges. Categorical variables are presented as percentages and counts within the complete observation set.'}

flextable::flextable(suppl_tbl$baseline_clusters_north) %>% 
  width(1, width = 3.5, unit = 'cm') %>% 
  width(2:4, width = 3.7, unit = 'cm') %>% 
  width(5, width = 2.6, unit = 'cm') %>% 
  width(6, width = 2.2, unit = 'cm') %>% 
  footnote(1, 1, 
           value = as_paragraph('BMI: body mass index, normal: BMI < 25 kg/m\u00B2, overweight: BMI 25 - 30 kg/m\u00B2, obesity: BMI > 30 kg/m\u00B2; Pre-CoV depression/anxiety: depression or anxiety before COVID-19; Freq. resp. infections: frequent (> 2 per year) respiratory infections; Freq. bact. Infections: frequent (> two per year) bacterial infections with antibiotic therapy; Pre-CoV sleep disorders: sleep disorders before COVID-19.'),
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 5:6, 
           value = as_paragraph('Categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic. Numeric variables: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic. P values corrected form multiple testing with Benjamini-Hochberg method.'),
           part = 'header', 
           ref_symbols = 'b') %>% 
  theme_vanilla

```{r tab-clust-base-it, tab.cap = 'Demographic and baseline clinical characteristic at the COVID-19 onset of the survey study participants assigned to the recovery clusters, Italy (IT) cohort. Numeric variables are presented as medians with interquartile ranges (IQR) and ranges. Categorical variables are presented as percentages and counts within the complete observation set.'}
